Cluster Analysis: Identifying Parkinson's Disease Subtypes

### 2015-07-16 writeup/abbreviated/abbreviated.tex
save figures for an abbreviated writeup

\documentclass[letterpaper,12pt]{article}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{numprint}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Random packages from
% http://tex.stackexchange.com/questions/50070/landscape-figure-in-latex
% Necessary for sideways pictures
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}
\usepackage{tablefootnote}
% for word wrap verbatim
\usepackage{listings}
\lstset{
   breaklines=true,
   basicstyle=\ttfamily}
% \pagestyle{fancy}
% \lhead{Jesse Mu}
% \rhead{CSCI339 Term Project}
\graphicspath{ {../figures/} }
% Or this, if run from main folder
% \graphicspath{ {./figures/} }

@@

\title{Cluster Analysis: Identifying Parkinson's Disease Subtypes}
\date{Wednesday, June 10}
\author{Jesse Mu}
\maketitle

@@

\subsection{Dimensionality Reduction: PCA}

May not be useful? If we're trying to identify \emph{clinically} relevant
features, merging them may not be a good idea. Regardless, Figure~\ref{fig:pca}
shows results of preliminary PCA.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{pca.pdf}
  \caption{PCA Analysis}
  \label{fig:pca}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{pca-eigenvalues.pdf}
  \caption{Scree test: eigenvalues by factor}
  \label{fig:pca-eigenvalues}
\end{figure}

Figure~\ref{fig:pca-eigenvalues} shows scree test elbow occurs around 2 or 2 or
.4 Also, eigenvalues $1-5 > 1$.

\clearpage
\section{$k$-means}
\subsection{Identifying optimal number of clusters}

\subsubsection{WSS Error Scree Test}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-wss-error.pdf}
  \caption{Scree test: WSS error by cluster size}
  \label{fig:kmeans-wss-error}
\end{figure}

Figure~\ref{fig:kmeans-wss-error} shows no optimal elbow in scree test! Maybe 2-3?

\subsubsection{Gap Statistic}

Optimal cluster is the local maximum of the gap statistic, but it appears to be
consistently increasing in Figure~\ref{fig:gap-statistic}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{gap-statistic.pdf}
  \caption{Gap statistic by cluster size}
  \label{fig:gap-statistic}
\end{figure}

\subsubsection{Average Silhouette Width}

Figure~\ref{fig:asw} shows average silhouette width as being consistently under
0.25 for all clusters, implying the data is not well structured.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{asw.pdf}
  \caption{Average silhouette width by cluster size}
  \label{fig:asw}
\end{figure}

% \subsubsection{\texttt{NbClust} package}

\subsection{Cluster statistics}
% CLUSTERS: 2
% ================================
% Sizes: 229 672
% WithinSS: 6117.986 7695.411
% Sum WithinSS: 13813.4
% CLUSTERS: 3
% ================================
% Sizes: 333 134 434
% WithinSS: 4669.037 4009.387 4153.892
% Sum WithinSS: 12832.32
% CLUSTERS: 4
% ================================
% Sizes: 79 394 275 153
% WithinSS: 2366.585 3356.709 3454.142 2879.508
% Sum WithinSS: 12056.94
\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l}
    $k$ & $n$ & Within SS & sum(Within SS) \\
    \hline
    2 & 229/672 & 6118/7695 & 13813 \\
    3 & 333/134/434 & 4669/40009/4154 & 12832 \\
    4 & 79/394/275/153 & 2367/3357/3454/2880  & 12057 \\
  \end{tabular}
  \caption{Cluster statistics}
  \label{tab:cluster-statistics}
\end{table}

\subsection{Silhouette plots}

Available in
Figures~\ref{fig:kmeans-silhouette-2},~\ref{fig:kmeans-silhouette-3},
and~\ref{fig:kmeans-silhouette-4}. Note: constructed with standardized
$z$-score data.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-silhouette-2.pdf}
  \caption{$k$-means cluster silhouette plot, $k = 2$}
  \label{fig:kmeans-silhouette-2}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-silhouette-3.pdf}
  \caption{$k$-means cluster silhouette plot, $k = 3$}
  \label{fig:kmeans-silhouette-3}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-silhouette-4.pdf}
  \caption{$k$-means cluster silhouette plot, $k = 4$}
  \label{fig:kmeans-silhouette-4}
\end{figure}

\subsection{Decision trees based on clusters}
% seed = 911
% CLUSTERS: 2
% ================================
% Complexity Parameter: 0.02183406
% 10-fold CV error: 0.1132075
% Root node error: 0.254162
% CLUSTERS: 3
% ================================
% Complexity Parameter: 0.01070664
% 10-fold CV error: 0.190899
% Root node error: 0.518313
% CLUSTERS: 4
% ================================
% Complexity Parameter: 0.01
% 10-fold CV error: 0.2552719
% Root node error: 0.5627081
\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l|l|l}
    $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
    validation} & Root Feature &
    Root Error & Figure \\
    \hline
    2 & 0.0218 & 0.113 & axial $\geq$ 4.5 & 0.254 & Figure~\ref{fig:kmeans-dtree-2} \\
    3 & 0.0107 & 0.191 & pigd $\geq$ 2.5 & 0.518 & Figure~\ref{fig:kmeans-dtree-3} \\
    4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
  \end{tabular}
  \caption{$k$-kmeans decision trees statistics}
  \label{tab:k-means-dtrees}
\end{table}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-2.pdf}
  \caption{Decision Tree from $k$-means clustering, 2 clusters}
  \label{fig:kmeans-dtree-2}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-3.pdf}
  \caption{Decision Tree from $k$-means clustering, 3 clusters}
  \label{fig:kmeans-dtree-3}
\end{figure}

@@

Available in
Figures~\ref{fig:kmeans-summaries-2},~\ref{fig:kmeans-summaries-3},
and~\ref{fig:kmeans-summaries-4}. Error bar is standard error.

@@

$k = 2$ seems too basic. Cluster is organized solely by severity - all
symptoms, including motor and nonmotor, are higher in severity in cluster 1,
and lower in cluster 2. Quite consistently, groups in cluster 1 are generally
of slightly higher age and pd duration.

$k = 3$ seems like a further development of $k = 2$, where clusters are simply
organized by linearly increasing severity.

$k = 4$ is where it gets interesting.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-summaries-2.pdf}
  \caption{Cluster Summaries, $k = 2$}
  \label{fig:kmeans-summaries-2}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{kmeans-summaries-3.pdf}
  \caption{Cluster Summaries, $k = 3$}
  \label{fig:kmeans-summaries-3}
\end{figure}

@@

\subsubsection{Ranked Features}

@@

\clearpage
\section{Affinity Propagation}

\subsection{Clustering}

Package \texttt{apcluster} was used. Distance matrix was the negative euclidean
squared distance ($r = 2$).

AP with input preferences minimized ($q = 0$) resulted in 8 clusters.
With the standard median input preferences ($q = 0.5$), algorithm failed to
converge with default parameters. Even setting damping factor to 0.98, maximum
iterations to 10000, and convergence iterations to 1000 failed to converge.
Might need to try a longer run.

\emph{However}, given that input preferences control how many clusters are
found, I don't think it's very useful to have some dozen clusters running
around.

\subsubsection{Silhouette Plots}

Silhouette plot in Figure~\ref{fig:ap-silhouette} looks pretty weak, really.
Tons of overlap between the clusters.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{ap-silhouette.pdf}
  \caption{AP silhouette plot, $k = 8$}
  \label{fig:ap-silhouette}
\end{figure}

\clearpage
\section{Hierarchical Clustering}

\subsection{Clustering}

Four dissimilarity methods were used with a euclidean distance matrix.
Dendrograms available in Figure~\ref{fig:hc-dendrograms}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{hc-dendrograms.pdf}
  \caption{Dendrograms}
  \label{fig:hc-dendrograms}
\end{figure}

\subsection{Cutting Trees}
\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l}
    Method & Condition & n & Figure \\
    \hline
    Complete & $k = 4$ & 4 (790/81/18/12) & ~\ref{fig:hc-summaries-complete-k4} \\
    Complete & \texttt{dynamicTreeCut}\tablefootnote{Package \texttt{dynamicTreeCut} in R (Langfelder P,
  Zhang B, Horvath S (2007)). Hybrid method, minimum cluster selection
parameters} & 13 (255/99/77/64/62/58/56/46/44/41/37/32/30) &
    ~\ref{fig:hc-summaries-complete-dynamic} \\
    Ward & $k = 4$ & 4 (200/237/263/201) & ~\ref{fig:hc-summaries-ward-D-k4} \\
    Ward & $h = 100$ & 3 (437/263/201) &
    ~\ref{fig:hc-summaries-ward-D-h100} \\
  \end{tabular}
  \caption{Clusters from Tree Cutting}
  \label{tab:tree-cutting}
\end{table}

\subsection{Interpretation}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{hc-summaries-complete-k4.pdf}
  \caption{Using maximum (complete linkage) dissimilarity, cutting tree for $k = 4$}
  \label{fig:hc-summaries-complete-k4}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{hc-summaries-complete-dynamic.pdf}
  \caption{Using maximum (complete linkage) dissimilarity, cutting tree
    dynamically}
  \label{fig:hc-summaries-complete-dynamic}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{hc-summaries-ward-D-k4.pdf}
  \caption{Using Ward (1963) dissimilarity, cutting tree for $k = 4$}
  \label{fig:hc-summaries-ward-D-k4}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{hc-summaries-ward-D-h100.pdf}
  \caption{Using Ward (1963) dissimilarity, cutting tree at $h = 100$}
  \label{fig:hc-summaries-ward-D-h100}
\end{figure}

\subsection{Interpretation}

Cluster sizes are available in Table~\ref{tab:ap-cluster-sizes}

\begin{table}[h]
  \centering
  \begin{tabular}{l|l}
  Cluster & Size \\
  \hline
  1 & 63 \\
  2 & 53 \\
  3 & 85 \\
  4 & 122 \\
  5 & 48 \\
  6 & 126 \\
  7 & 123 \\
  8 & 102 \\
  9 & 166 \\
  10 & 13 \\
  \end{tabular}
  \caption{AP Cluster Sizes}
  \label{tab:ap-cluster-sizes}
\end{table}

Boxplot summary of clusters available in Figure~\ref{fig:ap-summaries}.
\textbf{Discussion forthcoming.}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{ap-summaries.pdf}
  \caption{AP Boxplot Summaries}
  \label{fig:ap-summaries}
\end{figure}

\section{Biclustering}

Used BCBimax clustering algorithm. Clusters seem quite sparse.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{biclust-heatmaps-16.pdf}
  \caption{Biclustering heatmaps $N = 16$}
  \label{fig:biclust-heatmaps-16}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{biclust-bubbleplot-16.pdf}
  \caption{Bubbleplot $N = 16$}
  \label{fig:bubbleplot-16}
\end{figure}

\section{Subspace clustering}

% TODO: Wat.

\section{Bayesian Networks}

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{numprint}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Random packages from
% http://tex.stackexchange.com/questions/50070/landscape-figure-in-latex
% Necessary for sideways pictures
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}
\usepackage{tablefootnote}
% for word wrap verbatim
\usepackage{listings}
\lstset{
   breaklines=true,
   basicstyle=\ttfamily}
% \pagestyle{fancy}
% \lhead{Jesse Mu}
% \rhead{CSCI339 Term Project}
% One more .. since we're in the abbreviated subdir
\graphicspath{ {../../figures/abbreviated/} {../../figures/} }

@@

\title{Cluster Analysis: Identifying Parkinson's Disease Subtypes}
\maketitle

@@

\section{$k$-means}

$k$-means clustering with $k = 4$ was tried. $k = 2, 3$ provided models that
were too simplistic. $k = 5$ did not provide any new information, but rather
just fragmented existing groups.

\begin{table}[h]
  \centering
  \caption{Cluster statistics}
  \label{tab:cluster_stats}
  \begin{tabular}{l|l}
    Cluster & $n$ \\
    \hline
    1 & 79 \\
    2 & 394 \\
    3 & 275 \\
    4 & 153 \\
  \end{tabular}
\end{table}

\subsection{Decision tree}
\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l|l|l}
    $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
    validation} & Root Feature &
    Root Error & Figure \\
    \hline
    4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
  \end{tabular}
  \caption{$k$-kmeans decision trees statistics}
  \label{tab:k-means-dtrees}
\end{table}

@@

Available in Figure~\ref{fig:kmeans-summaries-4}. Error bar is standard error.

@@

@@

Post-hoc analysis using Tukey's HSD:
\begin{verbatim}
age insignificant differences:
         diff         lwr        upr        p adj
4-3 -2.271990  -4.7160927  0.1721117 7.920343e-02
sex insignificant differences:
           diff        lwr        upr     p adj
2-1 -0.09275204 -0.2454183 0.05991417 0.4000134
3-1 -0.14660529 -0.3046921 0.01148156 0.0803067
4-1  0.04757177 -0.1240051 0.21914864 0.8917054
3-2 -0.05385325 -0.1511666 0.04346014 0.4843788
pdonset insignificant differences:
           diff       lwr      upr     p adj
2-1 -0.90162565 -4.302019 2.498768 0.9037828
3-1 -0.05040276 -3.571532 3.470727 0.9999820
4-1 -0.53727145 -4.358869 3.284326 0.9837825
3-2  0.85122289 -1.316276 3.018722 0.7431486
4-2  0.36435420 -2.263251 2.991959 0.9844187
4-3 -0.48686869 -3.268952 2.295214 0.9695444
nms_d1 insignificant differences:
          diff         lwr       upr        p adj
3-2  0.5565021 -0.01586385  1.128868 6.018674e-02
nms_d3 insignificant differences:
          diff        lwr        upr       p adj
4-1  -1.192024  -4.493510   2.109461 0.789133802
nms_d4 insignificant differences:
          diff        lwr       upr        p adj
3-2  0.2951915 -0.3318339  0.922217 6.195406e-01
nms_d5 insignificant differences:
           diff         lwr       upr        p adj
3-2   0.7841071  -0.4814825  2.049697 0.3821565990
nms_d8 insignificant differences:
          diff       lwr      upr     p adj
4-1 -0.9010507 -2.846119 1.044017 0.6318563
3-2  0.7255838 -0.377602 1.828770 0.3280035
nms_d9 insignificant differences:
          diff         lwr       upr       p adj
4-1  0.7048896  -1.6510530  3.060832 0.867951568
tremor insignificant differences:
          diff       lwr        upr      p adj
3-1 -0.2662831 -1.064520 0.53195417 0.82613187
4-2 -0.5667198 -1.162396 0.02895624 0.06895758
rigidity insignificant differences:
          diff         lwr        upr        p adj
4-2  0.2310142 -0.03536748  0.4973959 1.153952e-01
\end{verbatim}

\subsubsection{Ranked Features by Information Gain}

@@

\subsubsection{Correlation Plots}

Figure~\ref{fig:corrplots}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{corrplots.pdf}
  \caption{Correlation plots}
  \label{fig:corrplots}
\end{figure}

\subsubsection{Bradykinesia and rigidity}

Figures~\ref{fig:bradykin-cisitot} and~\ref{fig:rigidity-cisitot}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{bradykin-cisitot.pdf}
  \caption{Relationship between bradykinesia and cisitot}
  \label{fig:bradykin-cisitot}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{bradykin-cisitot-condensed.pdf}
  \caption{Relationship between bradykinesia and cisitot (condensed)}
  \label{fig:bradykin-cisitot-condensed}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{bradykin-cisitot-all.pdf}
  \caption{Relationship between bradykinesia and cisitot (entire data)}
  \label{fig:bradykin-cisitot-all}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{rigidity-cisitot.pdf}
  \caption{Similar relationship between rigidity and cisitot}
  \label{fig:rigidity-cisitot}
\end{figure}

\section{Other Work}

\subsection{Bayesian Networks}

In Figure~\ref{fig:bnets}. Structure is too sparse, need to discretize or use
some kind of regularization (e.g. a lasso)

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\linewidth]{bnets.pdf}
  \caption{Bayesian Networks}
  \label{fig:bnets}
\end{figure}
